---
title: "Analysis of HC & T1D Single Cell Data"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries, mask functions, set paths, set ggplot theme

```{r}

library(Seurat)
library(dplyr)
library(ggplot2)
library(ComplexHeatmap)
library(circlize)
library(knitr)
library(stringr)
library(ggrepel)
library(tidyr)
library(igraph)

options(stringsAsFactors = FALSE)
set.seed(42)

`%notin%` <- Negate(`%in%`)

select   <- dplyr::select
rename   <- dplyr::rename
filter   <- dplyr::filter
distinct <- dplyr::distinct
mutate   <- dplyr::mutate

setwd("/Users/tybottorff/git/Bottorff_CD127_2026/HC_T1D_SINGLE_CELL/")

theme_minimal_white <- theme_bw(base_size = 14) + 
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white", color = "black"),
    strip.text = element_text(color = "black", size = 12),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    legend.background = element_rect(fill = "white", color = NA)
  )
theme_set(theme_minimal_white)

```

# Load in quality controlled data (see P348_4_HC_T1D_QC.Rmd)

```{r}

load("./saved_data/data10x_P348_4_norm_clusterd_corrected_wo_vdj.RData")

```

# Cluster cells by expression of selected surface proteins

```{r}

adt_markers_of_int <- c("CD27", "CD45RA", "CD45RO", "CD127 (IL-7R&#945;)","CD57 Recombinant","CD279","TIGIT (VSTM3)","KLRG1 (MAFA)")

DefaultAssay(data10x_P348_4_norm_clusterd_corrected_wo_vdj) <- "ADT"

# duplicate object, subsect ADT layer
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT <- data10x_P348_4_norm_clusterd_corrected_wo_vdj
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT[["ADT"]] <- subset(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT[["ADT"]], features = adt_markers_of_int)

# normalize, find features, scale, cluster
DefaultAssay(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT) <- "ADT"
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT <- NormalizeData(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT, normalization.method = 'CLR')
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT <- FindVariableFeatures(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT, selection.method = "vst")
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT <- ScaleData(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT <- RunPCA(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT, features = VariableFeatures(object = data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT))
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT <- FindNeighbors(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT, dims = 1:7)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT <- FindClusters(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT, resolution = c(0.4))
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT <- RunUMAP(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT, dims = 1:7)

# choose cluster numbers
Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT) <- "ADT_snn_res.0.4"
old_idents <- Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT)
new_idents <- as.character(old_idents)
new_idents[old_idents == "5"] <- "4"
new_idents[old_idents == "1"] <- "2"
new_idents[old_idents == "3"] <- "3"
new_idents[old_idents == "4"] <- "1"
new_idents[old_idents == "0"] <- "8"
new_idents[old_idents == "6"] <- "6"
new_idents[old_idents == "2"] <- "7"
new_idents[old_idents == "7"] <- "5"
Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT) <- factor(new_idents, levels = sort(unique(new_idents)))

```

# Display heatmap of clusters' average expression of selected surface proteins (figure 2A)

```{r}

avg_expr <- AverageExpression(
  data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT,
  layer = "scale.data"
)$ADT
# clean up row and column names
avg_expr_mat <- as.matrix(avg_expr)
rownames(avg_expr_mat) <- gsub(" .*", "", rownames(avg_expr_mat))   # e.g., "CD127 (IL-7Rα)" → "CD127"
colnames(avg_expr_mat) <- gsub("^g", "", colnames(avg_expr_mat))    # e.g., "g1" → "1"
# count number of cells per cluster
cluster_ids <- Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT)
cluster_sizes <- table(cluster_ids)
names(cluster_sizes) <- gsub("^g", "", names(cluster_sizes))
cluster_size_vector <- cluster_sizes[colnames(avg_expr_mat)]
cluster_bar_colors <- setNames(
  c("#302583", "#37773C", "#D9CB81", "#BF6E7A", "#95CAE9", "#61A899", "#9E4C95", "#7D2B54"),
  names(cluster_size_vector)
)
right_annotation <- rowAnnotation(
  "Cluster (C)\nsize\n(# cells)" = anno_barplot(
    matrix(cluster_size_vector, ncol = 1),
    bar_width = 0.8,
    gp = gpar(fill = cluster_bar_colors),
    border = FALSE
  ),
  annotation_name_side = "bottom"
)
fig2A <- pseudobulked_clustered_heatmap <- Heatmap(
  t(avg_expr_mat),
  name = "Z-score\nExpression",
  col = colorRamp2(c(-2, 0, 2), c("#45094e", "#b68b4d", "#fdf180")),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  right_annotation = right_annotation,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  column_names_gp = gpar(fontsize = 10),
  row_names_gp = gpar(fontsize = 10),
  row_order = c("1", "2", "3", "4", "5", "6", "7", "8"),
  column_order = c("CD45RA", "CD45RO", "CD127", "CD57", "CD279", "TIGIT", "KLRG1", "CD27"),
  border = TRUE
  )

fig2A

pdf("./figures/fig2A.pdf", width = 3, height = 2.5)
draw(pseudobulked_clustered_heatmap, heatmap_legend_side = "left")
dev.off()

```

# Look globally at all surface proteins' expression

```{r}

DefaultAssay(data10x_P348_4_norm_clusterd_corrected_wo_vdj) <- "ADT"
data10x_P348_4_norm_clusterd_corrected_wo_vdj <- FindVariableFeatures(data10x_P348_4_norm_clusterd_corrected_wo_vdj, selection.method = "vst", nfeatures = 2000)
data10x_P348_4_norm_clusterd_corrected_wo_vdj <- ScaleData(data10x_P348_4_norm_clusterd_corrected_wo_vdj)
data10x_P348_4_norm_clusterd_corrected_wo_vdj <- RunPCA(data10x_P348_4_norm_clusterd_corrected_wo_vdj, features = VariableFeatures(object = data10x_P348_4_norm_clusterd_corrected_wo_vdj))
data10x_P348_4_norm_clusterd_corrected_wo_vdj <- FindNeighbors(data10x_P348_4_norm_clusterd_corrected_wo_vdj, dims = 1:7)
data10x_P348_4_norm_clusterd_corrected_wo_vdj <- FindClusters(data10x_P348_4_norm_clusterd_corrected_wo_vdj)
data10x_P348_4_norm_clusterd_corrected_wo_vdj <- RunUMAP(data10x_P348_4_norm_clusterd_corrected_wo_vdj, dims = 1:7)
DefaultAssay(data10x_P348_4_norm_clusterd_corrected_wo_vdj) <- "RNA"

# assign cluster numbers again
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT$ADT_cluster <- 
  as.character(Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT))
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT$ADT_cluster <- 
  factor(
    data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT$ADT_cluster,
    levels = sort(unique(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT$ADT_cluster))
  )
data10x_P348_4_norm_clusterd_corrected_wo_vdj$ADT_cluster <- 
  as.character(Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj))
data10x_P348_4_norm_clusterd_corrected_wo_vdj$ADT_cluster <- 
  factor(
    data10x_P348_4_norm_clusterd_corrected_wo_vdj$ADT_cluster,
    levels = sort(unique(data10x_P348_4_norm_clusterd_corrected_wo_vdj$ADT_cluster))
  )

```

# Find differentially expressed transcripts and surface proteins between SKIM and non-SKIM (figure 2B, supplementary data 1)

```{r}

cluster_8_barcodes <- data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT@meta.data %>%
  filter(ADT_cluster == 8) %>%
  pull(cellBarcode)
data10x_P348_4_norm_clusterd_corrected_wo_vdj@meta.data$cluster_8_status <- ifelse(data10x_P348_4_norm_clusterd_corrected_wo_vdj@meta.data$cellBarcode %in% cluster_8_barcodes,
  "cluster-8",
  "not-cluster-8"
)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_pseudobulk <- AggregateExpression(
  data10x_P348_4_norm_clusterd_corrected_wo_vdj,
  assays = "ADT", 
  return.seurat = TRUE,
  group.by = c("sample_id", "cluster_8_status")
)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_pseudobulk$cluster_8_status <- factor(data10x_P348_4_norm_clusterd_corrected_wo_vdj_pseudobulk$cluster_8_status, levels = c("cluster-8", "not-cluster-8"))
Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_pseudobulk) <- "cluster_8_status"
pseudobulk_degs <- FindMarkers(
  object = data10x_P348_4_norm_clusterd_corrected_wo_vdj_pseudobulk,
  ident.1 = "cluster-8",
  ident.2 = "not-cluster-8",
  assay = "ADT",
  test.use = "MAST",
  latent.vars = "sample_id"
)
pseudobulk_degs$gene <- rownames(pseudobulk_degs)
volcano_df <- pseudobulk_degs %>%
  mutate(
    gene = case_when(
      str_detect(gene, "^TCR") ~ "TCR Vα7.2",
      TRUE ~ word(gene, 1)  # keep only the first word otherwise
    ),
    label_gene = ifelse(p_val_adj < 0.05 & abs(avg_log2FC) > 1, gene, NA))

volcano_df$significance <- with(volcano_df, ifelse(p_val_adj < 0.05 & avg_log2FC > 0, "Up in SKIM",
                                    ifelse(p_val_adj < 0.05 & avg_log2FC < 0, "Down in SKIM", "NS")))
volcano_df_adt <- volcano_df %>%
  mutate(modality = "surface proteins")
# RNA layer now
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT@meta.data$cluster_8_status <- ifelse(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT@meta.data$cellBarcode %in% cluster_8_barcodes,
  "cluster-8",
  "not-cluster-8"
)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_pseudobulk <- AggregateExpression(
  data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT,
  assays = "RNA", 
  return.seurat = TRUE,
  group.by = c("sample_id", "cluster_8_status")
)
data10x_P348_4_norm_clusterd_corrected_wo_vdj@meta.data$cluster_8_status <- ifelse(data10x_P348_4_norm_clusterd_corrected_wo_vdj@meta.data$cellBarcode %in% cluster_8_barcodes,
  "cluster-8",
  "not-cluster-8"
)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_pseudobulk$cluster_8_status <- factor(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_pseudobulk$cluster_8_status, levels = c("cluster-8", "not-cluster-8"))
Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_pseudobulk) <- "cluster_8_status"
Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_pseudobulk)
pseudobulk_degs <- FindMarkers(
  object = data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_pseudobulk,
  ident.1 = "cluster-8",
  ident.2 = "not-cluster-8",
  assay = "RNA",
  test.use = "DESeq2"
)
pseudobulk_degs$gene <- rownames(pseudobulk_degs)
volcano_df <- pseudobulk_degs %>%
  mutate(label_gene = ifelse(p_val_adj < 0.05 & abs(avg_log2FC) > 1, gene, NA))
volcano_df$significance <- with(volcano_df, ifelse(p_val_adj < 0.05 & avg_log2FC > 0, "Up in SKIM",
                                    ifelse(p_val_adj < 0.05 & avg_log2FC < 0, "Down in SKIM", "NS")))
volcano_df_rna <- volcano_df %>%
  mutate(modality = "transcripts")
volcano_df <- bind_rows(volcano_df_adt, volcano_df_rna)
fig2B <- ggplot(volcano_df %>% filter(!(is.na(significance))), aes(x = avg_log2FC, y = -log10(p_val_adj), color = significance)) +
  geom_point(size = 1.25, alpha = 0.75) +
  geom_text_repel(aes(label = label_gene), size = 6, na.rm = TRUE, max.overlaps = 4) +
  scale_color_manual(values = c("Up in SKIM" = "#7D2B54", "Down in SKIM" = "black", "NS" = "grey")) +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 adjusted\np-value",
    color = "Significance"
  ) +
  facet_wrap(~ modality, scales = "free") +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 14),
        legend.key = element_blank()) +  # Remove the key (color boxes)
  guides(color = guide_legend(override.aes = list(size = 2)))  # Ensure that only color legend is shown
fig2B
ggsave("./figures/fig2B.pdf", width = 8.4, height = 3.25)
write.csv(volcano_df, "./saved_data/supp_data_1.csv", row.names = FALSE)

```

# Bring in TCRseq data

```{r}

tcrs <- read.csv("./raw_data/P348-4_aggr_vdj_filtered_contig_annotations.csv") %>%
  # filter for cells with no more than 1 alpha or beta chain (<= 1 each)
  group_by(barcode, chain) %>%
  mutate(per_cell_chain_count = n()) %>% 
  filter(per_cell_chain_count == 1) %>%
  ungroup() %>%
  group_by(barcode) %>% 
  mutate(TRA_TRB_pair = n()) %>% 
  filter(TRA_TRB_pair == 2) %>%
  ungroup() %>%
  rename(cellBarcode = barcode,
         studyGroup = Meta.studyGroup) %>%
  pivot_wider(id_cols = c(cellBarcode, studyGroup), names_from = "chain", values_from = c("v_gene", "d_gene", "j_gene", "c_gene", "cdr3", "cdr3_nt"))
adt_tcr_joined_data <- data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT@meta.data %>%
  # subset for cells in both scRNAseq and scTCRseq (inner_join)
  inner_join(tcrs, by = c("cellBarcode", "studyGroup"))

```

# Plot TCRvalpha7.2+ (%) by cluster (figure 2C)

```{r}

mait_perc_plot_df <- adt_tcr_joined_data %>%
  select(cellBarcode, sampleID, studyGroup, ADT_cluster, v_gene_TRA, j_gene_TRA) %>%
  group_by(ADT_cluster, sampleID) %>%
  summarize(
    # sum cells with identifiable v/j genes for TRA
    total_cells_with_info = sum(!is.na(v_gene_TRA) & !is.na(j_gene_TRA)),
    MAIT_cells = sum(v_gene_TRA == "TRAV1-2" & j_gene_TRA %in% c("TRAJ33", "TRAJ12", "TRAJ20"))) %>%
  mutate(
    MAIT_percentage = (MAIT_cells / total_cells_with_info) * 100,
    cluster_proportion = total_cells_with_info / sum(total_cells_with_info),
    weighted_MAIT_contribution = MAIT_percentage * cluster_proportion
  )
fig2C <- ggplot(mait_perc_plot_df, aes(x = as.factor(ADT_cluster), y = MAIT_percentage, fill = as.factor(ADT_cluster))) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(
    values = c(
      "1" = "#302583",
      "2" = "#37773C",
      "3" = "#D9CB81",
      "4" = "#BF6E7A",
      "5" = "#95CAE9",
      "6" = "#61A899",
      "7" = "#9E4C95",
      "8" = "#7D2B54"
    )) +
  geom_jitter(size = 1) +
  labs(
    x = "Cluster",
    y = "TCRvalpha7.2+ (%)",
    fill = "Cluster"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
        legend.position = "none") +
  ylim(0, 100)
fig2C
ggsave("./figures/fig2C.pdf", width = 4, height = 2.75)

```

# Find differentially expressed transcripts and surface proteins between non-MAIT SKIM and non-SKIM (figure 2D, supplementary data 2)

```{r}

# subset for TCRseq-containing data in cluster 8
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_with_tcr_data_cluster_8 <- subset(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT, subset = cellBarcode %in% tcrs$cellBarcode & ADT_cluster == 8)
# define MAIT cells as those with alpha chains expressing TRAV1-2 and one of the following: TRAJ12, TRAJ20, or TRAJ33
maits_cells <- tcrs %>%
  filter(v_gene_TRA == "TRAV1-2" & j_gene_TRA %in% c("TRAJ33", "TRAJ12", "TRAJ20")) %>%
  pull(cellBarcode)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_with_tcr_data_cluster_8@meta.data$mait_status <- ifelse(
  data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_with_tcr_data_cluster_8@meta.data$cellBarcode %in% maits_cells,
  "MAIT",
  "not-MAIT"
)
cluster_8_mait_or_tcr_neg_barcodes <- data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT@meta.data %>%
  filter(ADT_cluster == 8,
         cellBarcode %in% maits_cells | !(cellBarcode %in% tcrs$cellBarcode)) %>%
  pull(cellBarcode)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_no_cluster_8_maits <- subset(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT, subset = !(cellBarcode %in% cluster_8_mait_or_tcr_neg_barcodes))
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_no_cluster_8_maits_pseudobulk <- AggregateExpression(
  data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_no_cluster_8_maits,
  assays = "RNA", 
  return.seurat = TRUE,
  group.by = c("sample_id", "cluster_8_status")
)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_no_cluster_8_maits_pseudobulk$cluster_8_status <- factor(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_no_cluster_8_maits_pseudobulk$cluster_8_status, levels = c("cluster-8", "not-cluster-8"))
Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_no_cluster_8_maits_pseudobulk) <- "cluster_8_status"
pseudobulk_degs <- FindMarkers(
  object = data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT_no_cluster_8_maits_pseudobulk,
  ident.1 = "cluster-8",
  ident.2 = "not-cluster-8",
  assay = "RNA",
  test.use = "DESeq2"
)
pseudobulk_degs$gene <- rownames(pseudobulk_degs)
volcano_df <- pseudobulk_degs %>%
  mutate(label_gene = ifelse(p_val_adj < 0.05 & abs(avg_log2FC) > 1, gene, NA))
volcano_df$significance <- with(volcano_df, ifelse(p_val_adj < 0.05 & avg_log2FC > 0, "Up in SKIM\nnon-MAITs",
                                    ifelse(p_val_adj < 0.05 & avg_log2FC < 0, "Down in SKIM\nnon-MAITs", "NS")))
volcano_df_rna <- volcano_df %>%
  mutate(modality = "transcripts")
data10x_P348_4_norm_clusterd_corrected_wo_vdj_no_cluster_8_maits <- subset(data10x_P348_4_norm_clusterd_corrected_wo_vdj, subset = !(cellBarcode %in% cluster_8_mait_or_tcr_neg_barcodes))
data10x_P348_4_norm_clusterd_corrected_wo_vdj_no_cluster_8_maits_pseudobulk <- AggregateExpression(
  data10x_P348_4_norm_clusterd_corrected_wo_vdj_no_cluster_8_maits,
  assays = "ADT", 
  return.seurat = TRUE,
  group.by = c("sample_id", "cluster_8_status")
)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_no_cluster_8_maits_pseudobulk$cluster_8_status <- factor(data10x_P348_4_norm_clusterd_corrected_wo_vdj_no_cluster_8_maits_pseudobulk$cluster_8_status, levels = c("cluster-8", "not-cluster-8"))
Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_no_cluster_8_maits_pseudobulk) <- "cluster_8_status"
pseudobulk_degs <- FindMarkers(
  object = data10x_P348_4_norm_clusterd_corrected_wo_vdj_no_cluster_8_maits_pseudobulk,
  ident.1 = "cluster-8",
  ident.2 = "not-cluster-8",
  assay = "ADT",
  test.use = "MAST"
)
pseudobulk_degs$gene <- rownames(pseudobulk_degs)
volcano_df <- pseudobulk_degs %>%
  mutate(
    gene = case_when(
      str_detect(gene, "^TCR") ~ "TCR Vα7.2",
      TRUE ~ word(gene, 1)  # keep only the first word otherwise
    ),
    label_gene = ifelse(p_val_adj < 0.05 & abs(avg_log2FC) > 1, gene, NA))
volcano_df$significance <- with(volcano_df, ifelse(p_val_adj < 0.05 & avg_log2FC > 0, "Up in SKIM\nnon-MAITs",
                                    ifelse(p_val_adj < 0.05 & avg_log2FC < 0, "Down in SKIM\nnon-MAITs", "NS")))
volcano_df_adt <- volcano_df %>%
  mutate(modality = "surface proteins")
volcano_df <- bind_rows(volcano_df_adt, volcano_df_rna)
fig2D <- ggplot(volcano_df %>% filter(!(is.na(significance))), aes(x = avg_log2FC, y = -log10(p_val_adj), color = significance)) +
  geom_point(size = 1.25, alpha = 0.75) +
  geom_text_repel(aes(label = label_gene), size = 4.5, na.rm = TRUE, max.overlaps = Inf) +
  scale_color_manual(values = c("Up in SKIM\nnon-MAITs" = "#7D2B54", "Down in SKIM\nnon-MAITs" = "black", "NS" = "grey")) +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 adjusted\np-value",
    color = "Significance"
  ) +
  facet_wrap(~ modality, scales = "free") +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 13),
        legend.key = element_blank(),
        axis.text.x = element_text(size = 11),
        axis.text.y = element_text(size = 11),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        strip.text = element_text(size = 14)) +
  guides(color = guide_legend(override.aes = list(size = 2)))
fig2D
ggsave("./figures/fig2D.pdf", width = 6.4, height = 2.4)
write.csv(volcano_df, "./saved_data/supp_data_2.csv", row.names = FALSE)

```

# Find differentially expressed transcripts and surface proteins between CD127+ Tex and CD127- Tex (figure 6C, supplementary data 4)

```{r}
# CD127+ Tex (cluster 4) was old cluster 5, CD57+ Tex (1) was old 4, PD1+ Tex (2) was old 1
cluster_1_4_5_barcodes <- data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT@meta.data %>%
  filter(ADT_snn_res.0.4 %in% c(1, 4, 5)) %>%
  pull(cellBarcode)
cluster_5_barcodes <- data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT@meta.data %>%
  filter(ADT_snn_res.0.4 == 5) %>%
  pull(cellBarcode)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5 <- subset(data10x_P348_4_norm_clusterd_corrected_wo_vdj, subset = cellBarcode %in% cluster_1_4_5_barcodes)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5@meta.data$cluster_5_status <- ifelse(
data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5@meta.data$cellBarcode %in% cluster_5_barcodes,
  "cluster-5",
  "not-cluster-5"
)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5_pseudobulk <- AggregateExpression(
  data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5,
  assays = "RNA", 
  return.seurat = TRUE,
  group.by = c("sample_id", "cluster_5_status")
)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5_pseudobulk$cluster_5_status <- factor(data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5_pseudobulk$cluster_5_status, levels = c("cluster-5", "not-cluster-5"))
Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5_pseudobulk) <- "cluster_5_status"
pseudobulk_degs <- FindMarkers(
  object = data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5_pseudobulk,
  ident.1 = "cluster-5",
  ident.2 = "not-cluster-5",
  assay = "RNA",
  test.use = "DESeq2"
)
pseudobulk_degs$gene <- rownames(pseudobulk_degs)
volcano_df <- pseudobulk_degs %>%
  mutate(label_gene = ifelse(p_val_adj < 0.05 & abs(avg_log2FC) > 0.75, gene, NA))
volcano_df$significance <- with(volcano_df,
  ifelse(is.na(p_val_adj) | p_val_adj >= 0.05, "NS",
    ifelse(avg_log2FC > 0, "Up in CD127+ Tex", "Up in CD127- Tex")
  )
)
volcano_df_rna <- volcano_df %>%
  mutate(modality = "transcripts")
data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5_pseudobulk <- AggregateExpression(
  data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5,
  assays = "ADT", 
  return.seurat = TRUE,
  group.by = c("sample_id", "cluster_5_status")
)
data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5_pseudobulk$cluster_5_status <- factor(data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5_pseudobulk$cluster_5_status, levels = c("cluster-5", "not-cluster-5"))
Idents(data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5_pseudobulk) <- "cluster_5_status"
pseudobulk_degs <- FindMarkers(
  object = data10x_P348_4_norm_clusterd_corrected_wo_vdj_clusters_1_4_5_pseudobulk,
  ident.1 = "cluster-5",
  ident.2 = "not-cluster-5",
  assay = "ADT",
  test.use = "MAST"
)
pseudobulk_degs$gene <- rownames(pseudobulk_degs)
volcano_df <- pseudobulk_degs %>%
  mutate(label_gene = ifelse(p_val_adj < 0.05 & abs(avg_log2FC) > 0.75, gene, NA))
volcano_df$significance <- with(volcano_df,
  ifelse(is.na(p_val_adj) | p_val_adj >= 0.05, "NS",
    ifelse(avg_log2FC > 0, "Up in CD127+ Tex", "Up in CD127- Tex")
  )
)
volcano_df_adt <- volcano_df %>%
  mutate(modality = "surface proteins")
volcano_df <- bind_rows(volcano_df_rna, volcano_df_adt) %>%
  mutate(label_gene = gsub(" .*", "", label_gene),
         significance = case_when(
           significance == "Up in CD127+ Tex" ~ "Up in\nCD127+ Tex",
           significance == "Up in CD127- Tex" ~ "Up in\nCD127- Tex",
           TRUE ~ significance
         ))
fig6C <- ggplot(volcano_df %>% filter(!(is.na(significance))), aes(x = avg_log2FC, y = -log10(p_val_adj), color = significance)) +
  geom_point(size = 1.25, alpha = 0.75) +
  geom_text_repel(aes(label = label_gene), size = 6, na.rm = TRUE, max.overlaps = 10) +
  scale_color_manual(values = c("Up in\nCD127+ Tex" = "#BF6E7A", "Up in\nCD127- Tex" = "black", "NS" = "grey")) +
  labs(
    x = "Log2 Fold Change",
    y = "-Log10 adjusted\np-value",
    color = "Significance"
  ) +
  facet_wrap(~ modality, scales = "free") +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 14),
        legend.key = element_blank()) +
  guides(color = guide_legend(override.aes = list(size = 2)))
fig6C
ggsave("./figures/fig6C.pdf", fig6C, width = 7.5, height = 3)
write.csv(volcano_df, "./saved_data/supp_data_4.csv", row.names = FALSE)

```

# Calculate Jaccard similarities between clusters' TCR repertoires (figure 7D)

```{r}

# The Jaccard similarity index = number observations in both sets / number observations in either set (so intersection / union)
clonotype_df <- adt_tcr_joined_data %>%
  mutate(mait_status = case_when(
   v_gene_TRA == "TRAV1-2" & j_gene_TRA %in% c("TRAJ33", "TRAJ12", "TRAJ20") ~ "MAIT",
   is.na(v_gene_TRA) | is.na(j_gene_TRA) ~ "unidentifiable",
   !is.na(v_gene_TRA) & v_gene_TRA != "TRAV1-2" | !is.na(j_gene_TRA) & !(j_gene_TRA %in% c("TRAJ33", "TRAJ12", "TRAJ20")) ~ "not MAIT"
  )) %>%
  mutate(clonotype = paste(v_gene_TRA, j_gene_TRA, cdr3_nt_TRA, v_gene_TRB, j_gene_TRB, cdr3_nt_TRB, sep = "_")) %>%
  select(cellBarcode, sampleID, ADT_cluster, clonotype, mait_status, v_gene_TRA, j_gene_TRA)
# downsample clusters so each has same number of TCRs
set.seed(123)
tcr_counts <- clonotype_df %>%
  group_by(ADT_cluster) %>%
  summarize(n = n())
min_tcrs <- min(tcr_counts$n)
clonotype_df <- clonotype_df %>%
  group_by(ADT_cluster) %>%
  slice_sample(n = min_tcrs) %>%
  ungroup()
# define function to calculate Jaccard index
calculate_jaccard_similarity <- function(df, cluster_col, clonotype_col, cluster1, cluster2) {
  cluster1_data <- df[df[[cluster_col]] == cluster1, ]
  cluster2_data <- df[df[[cluster_col]] == cluster2, ]
  clonotypes_cluster1 <- unique(cluster1_data[[clonotype_col]])
  clonotypes_cluster2 <- unique(cluster2_data[[clonotype_col]])
  intersection <- length(intersect(clonotypes_cluster1, clonotypes_cluster2))
  union <- length(union(clonotypes_cluster1, clonotypes_cluster2))
  jaccard_index <- intersection / union
  return(jaccard_index)
}
# calculate Jaccard indices across all cluster comparisons
calculate_pairwise_jaccard <- function(df, cluster_col, clonotype_col) {
  clusters <- unique(df[[cluster_col]])
  jaccard_matrix <- matrix(0, nrow = length(clusters), ncol = length(clusters))
  rownames(jaccard_matrix) <- clusters
  colnames(jaccard_matrix) <- clusters
  for (i in 1:length(clusters)) {
    for (j in i:length(clusters)) {
      jaccard_value <- calculate_jaccard_similarity(df, cluster_col, clonotype_col, clusters[i], clusters[j])
      jaccard_matrix[i, j] <- jaccard_value
      jaccard_matrix[j, i] <- jaccard_value
    }
  }
  return(jaccard_matrix)
}
jaccard_matrix <- calculate_pairwise_jaccard(clonotype_df, "ADT_cluster", "clonotype")
new_order <- c("4", "1", "2", "3", "5", "6", "7", "8")
jaccard_matrix <- jaccard_matrix[new_order, new_order]
jaccard_matrix[upper.tri(jaccard_matrix, diag = TRUE)] <- NA
# scale color
max_jaccard_value <- max(jaccard_matrix, na.rm = TRUE)
col_fun <- colorRamp2(c(0, max_jaccard_value), c("white", "#6A0DAD"))
fig7D <- Heatmap(
    jaccard_matrix,
    name = "TCR repertoire overlap",
    col = col_fun,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_dend_side = "left",
    column_dend_side = "top",
    show_row_dend = FALSE,
    show_column_dend = FALSE,
    row_names_side = "left",
    column_names_side = "top",
    na_col = "grey")

fig7D

pdf("./figures/fig7D.pdf", width = 5, height = 3)
draw(fig7D)
dev.off()

```

# Combine clusters 5-7 and summarize sharing in a node summary (figure 7E)
 - scale nodes by the number of rare mitochondrial variants

```{r}

clonotype_df <- adt_tcr_joined_data %>%
  mutate(mait_status = case_when(
   v_gene_TRA == "TRAV1-2" & j_gene_TRA %in% c("TRAJ33", "TRAJ12", "TRAJ20") ~ "MAIT",
   is.na(v_gene_TRA) | is.na(j_gene_TRA) ~ "unidentifiable",
   !is.na(v_gene_TRA) & v_gene_TRA != "TRAV1-2" | !is.na(j_gene_TRA) & !(j_gene_TRA %in% c("TRAJ33", "TRAJ12", "TRAJ20")) ~ "not MAIT"
  )) %>%
  mutate(clonotype = paste(v_gene_TRA, j_gene_TRA, cdr3_nt_TRA, v_gene_TRB, j_gene_TRB, cdr3_nt_TRB, sep = "_")) %>%
  select(cellBarcode, sampleID, ADT_cluster, clonotype, mait_status, v_gene_TRA, j_gene_TRA)
# downsample clusters so each has same number of TCRs
set.seed(123)
tcr_counts <- clonotype_df %>%
  group_by(ADT_cluster) %>%
  summarize(n = n())
min_tcrs <- min(tcr_counts$n)
clonotype_df <- clonotype_df %>%
  group_by(ADT_cluster) %>%
  slice_sample(n = min_tcrs) %>%
  ungroup()
calculate_pairwise_jaccard <- function(df, cluster_col, clonotype_col) {
  clusters <- unique(df[[cluster_col]])
  jaccard_matrix <- matrix(0, nrow = length(clusters), ncol = length(clusters))
  rownames(jaccard_matrix) <- clusters
  colnames(jaccard_matrix) <- clusters
  for (i in 1:length(clusters)) {
    for (j in i:length(clusters)) {
      jaccard_value <- calculate_jaccard_similarity(df, cluster_col, clonotype_col, clusters[i], clusters[j])
      jaccard_matrix[i, j] <- jaccard_value
      jaccard_matrix[j, i] <- jaccard_value
    }
  }
  return(jaccard_matrix)
}
jaccard_matrix <- calculate_pairwise_jaccard(clonotype_df, "ADT_cluster", "clonotype")
# create mapping for cluster collapse
cluster_map <- c("1"="1", "2"="2", "3"="3", "4"="4", "5"="x", "6"="x", "7"="x", "8"="8")
jaccard_collapsed <- matrix(NA, nrow=6, ncol=6)
rownames(jaccard_collapsed) <- colnames(jaccard_collapsed) <- c("1", "2", "3", "4", "x", "8")
unique_clusters <- rownames(jaccard_collapsed)
for (i in unique_clusters) {
  for (j in unique_clusters) {
    # Find original cluster labels that map to these
    rows_i <- names(cluster_map)[cluster_map == i]
    cols_j <- names(cluster_map)[cluster_map == j]

    # Extract relevant submatrix and compute mean (excluding NAs)
    sub_values <- jaccard_matrix[rows_i, cols_j]
    jaccard_collapsed[i, j] <- mean(sub_values, na.rm = TRUE)
  }
}
jaccard_long <- as.data.frame(as.table(jaccard_collapsed)) %>%
  rename(from = Var1, to = Var2, jaccard = Freq) %>%
  mutate(from = as.character(from), to = as.character(to))
# get between-cluster edges
jaccard_between <- jaccard_long %>%
  filter(from != to) %>%
  mutate(key = paste(pmin(from, to), pmax(from, to), sep = "-")) %>%
  distinct(key, .keep_all = TRUE)
# count number of repeated clonotypes per cluster
tcr_expansion <- clonotype_df %>%
  mutate(ADT_cluster = case_when(
    ADT_cluster %in% c("5", "6", "7") ~ "x",
    TRUE ~ ADT_cluster
  )) %>%
  group_by(ADT_cluster, clonotype) %>%
  filter(n() > 1) %>%
  summarise(n_repeats = n(), .groups = "drop") %>%
  group_by(ADT_cluster) %>%
  summarise(expansion_score = sum(n_repeats), .groups = "drop") %>%
  rename(from = ADT_cluster)
self_edges <- tcr_expansion %>%
  mutate(to = from, jaccard = NA) %>%
  select(from, to, jaccard, expansion_score)
# for between-cluster edges, jaccard drives edge width
edges_between <- jaccard_between %>%
  select(from, to, jaccard) %>%
  mutate(expansion_score = NA)
# for self-loops, expansion_score drives width
edges_self <- self_edges
# combine edge list
edges_final <- bind_rows(edges_between, edges_self)
match_TCR_donor_all_cluster_donor_tcr_count <- clonotype_df %>%
  mutate(ADT_cluster = case_when(
    ADT_cluster %in% c("5", "6", "7") ~ "x",
    TRUE ~ ADT_cluster
  )) %>%
  group_by(cluster = ADT_cluster, sampleID, clonotype) %>%
  summarise(n_cells = n()) %>%
  filter(n_cells > 1) %>%
  group_by(cluster) %>%
  summarise(number_cells_with_tcr = sum(n_cells), .groups = "drop") %>%
  rename(from = cluster)
vertices <- match_TCR_donor_all_cluster_donor_tcr_count %>%
  mutate(from = as.character(from)) %>%
  distinct(from, number_cells_with_tcr) %>%
  rename(name = from)
network_graph <- graph_from_data_frame(
  d = edges_final,
  vertices = vertices,
  directed = FALSE
)
cluster_pal <- c(
  "1" = "#302583",
  "2" = "#37773C",
  "3" = "#D9CB81",
  "4" = "#BF6E7A",
  "x" = "grey",
  "8" = "#7D2B54"
)
V(network_graph)$color <- cluster_pal[V(network_graph)$name]
# scale nodes by the mean number of rare mitochondrial variants, when possible (otherwise use 2)
V(network_graph)$size <- c(5.6, 3.5, 2, 1.44, 2, 1.62)*6
eps <- 1e-6
E(network_graph)$width <- ifelse(
  is.na(E(network_graph)$expansion_score),
  scales::rescale(E(network_graph)$jaccard + eps, to = c(0.5, 8)),
  scales::rescale(E(network_graph)$expansion_score + eps, to = c(0.5, 8))
)
E(network_graph)$curved <- ifelse(
  ends(network_graph, E(network_graph))[,1] == ends(network_graph, E(network_graph))[,2],
  0.2,
  0.05
)
n_nodes <- vcount(network_graph)
grid_layout <- matrix(nrow = n_nodes, ncol = 2)
grid_layout[,1] <- rep(1:3, length.out = n_nodes) * 1
grid_layout[,2] <- rep(1:3, each = ceiling(n_nodes / 3))[1:n_nodes] * 1
grid_layout <- grid_layout * 0.6
x_range <- range(grid_layout[,1])
y_range <- range(grid_layout[,2])
x_padding <- diff(x_range) * 0.5
y_padding <- diff(y_range) * 0.5
pdf("./figures/fig7C.pdf", width = 8, height = 8)
plot(
  network_graph,
  layout = grid_layout,
  rescale = FALSE,
  xlim = x_range + c(-x_padding, x_padding),
  ylim = y_range + c(-y_padding, y_padding),
  vertex.label.font = 2,
  vertex.label.color = "black",
  vertex.label.cex = 1.5,
  edge.color = "black"
)
dev.off()

```

# Calculate expression of public Tc17 module in each cluster (figure 2F)

```{r}

# gene list comes from supplementary table 2 from PMID 34914499
tc17 = c("RORC", "ZBTB16", "CEBPD", "RORA", "NR1D1", "CCL20", "CD40LG", "LTB", "TNFSF13B", "IL26", "IL17A", "FLT3LG", "TNF", "IL23A", "CCR6", "IL23R", "IL7R", "IL17RE", "IL18RAP", "SLC4A10", "KLRB1", "TMIGD2", "IL4I1", "NCR3", "LTK", "CA2", "ME1", "AQP3", "ADAM12")
DefaultAssay(data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT) <- "RNA"
expr_mat <- GetAssayData(
  data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT,
  layer = "data"
)
genes_use <- intersect(tc17, rownames(expr_mat))
meta_rna <- data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT@meta.data %>%
  dplyr::select(sampleID, cluster = ADT_snn_res.0.4) %>%
  mutate(cluster = case_when(
    cluster == "5" ~ "4",
    cluster == "1" ~ "2",
    cluster == "4" ~ "1",
    cluster == "0" ~ "8",
    cluster == "2" ~ "7",
    cluster == "7" ~ "5",
    TRUE ~ cluster
  ))
pb_rna <- meta_rna %>%
  cbind(t(expr_mat[genes_use, , drop = FALSE])) %>%
  pivot_longer(cols = all_of(genes_use), names_to = "gene", values_to = "expr") %>%
  group_by(sampleID, cluster) %>%
  summarise(module_expr = mean(expr), .groups = "drop")
cluster_cols <- c(
  "1"="#302583","2"="#37773C","3"="#D9CB81","4"="#BF6E7A",
  "5"="#95CAE9","6"="#61A899","7"="#9E4C95","8"="#7D2B54"
)
fig2F <- ggplot(pb_rna, aes(x=factor(cluster), y=module_expr, fill=factor(cluster))) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 0.6) +
  scale_fill_manual(values = cluster_cols) +
  labs(
    x = "Cluster",
    y = "Mean RNA\nexpression\nof Tc17 module"
  )
fig2F
ggsave(
  "./figures/fig2F.pdf",
  width = 6,
  height = 2.5
)
```

# Calculate expression of public terminal Tex module in each cluster (figure 6A)

```{r}

# gene list comes from supplementary table 2 from PMID 34914499
terminal_tex = c("RBPJ", "ETV1", "TOX", "ZBED2", "TOX2", "CXCL13", "TNFSF4", "FAM3C", "GZMB", "CSF1", "CCL3", "CD70", "IFNG", "NAMPT", "FASLG", "IL2RA", "CXCR6", "CD74", "IL2RB", "IL2RG", "TNFRSF9", "LAYN", "ENTPD1", "HAVCR2", "CTLA4", "KRT86", "TNFRSF18", "GEM", "TIGIT", "DUSP4")
genes_use <- intersect(terminal_tex, rownames(expr_mat))
meta_rna <- data10x_P348_4_norm_clusterd_corrected_wo_vdj_subset_ADT@meta.data %>%
  dplyr::select(sampleID, cluster = ADT_snn_res.0.4) %>%
  mutate(cluster = case_when(
    cluster == "5" ~ "4",
    cluster == "1" ~ "2",
    cluster == "4" ~ "1",
    cluster == "0" ~ "8",
    cluster == "2" ~ "7",
    cluster == "7" ~ "5",
    TRUE ~ cluster
  ))
pb_rna <- meta_rna %>%
  cbind(t(expr_mat[genes_use, , drop = FALSE])) %>%
  pivot_longer(cols = all_of(genes_use), names_to = "gene", values_to = "expr") %>%
  group_by(sampleID, cluster) %>%
  summarise(module_expr = mean(expr), .groups = "drop")
cluster_cols <- c(
  "1"="#302583","2"="#37773C","3"="#D9CB81","4"="#BF6E7A",
  "5"="#95CAE9","6"="#61A899","7"="#9E4C95","8"="#7D2B54"
)
fig6A <- ggplot(pb_rna, aes(x=factor(cluster), y=module_expr, fill=factor(cluster))) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 0.6) +
  scale_fill_manual(values = cluster_cols) +
  labs(
    x = "Cluster",
    y = "Mean RNA\nexpression\nof terminal\nTex module"
  )
fig6A
ggsave(
  "./figures/fig6A.pdf",
  width = 6.5,
  height = 2.5
)

```