---
title: "Analysis of bulk ATAC-sequencing of non-naive CD8s from AbATE"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries, set theme and path

```{r}

library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(ggplot2)
library(ggthemes)
library(ggbeeswarm)
library(viridis)
library(stringr)
library(kableExtra)
library(RColorBrewer)
library(ggrepel)
library(ggpubr)
library(plotly)
library(GenomicRanges)
library(GenomicAlignments)
library(EnsDb.Hsapiens.v86)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
library(ChIPpeakAnno)
library(DiffBind)
library(ATACseqQC)
library(soGGi)
library(chromVAR)
library(SummarizedExperiment)
library(edgeR)
library(limma)
library(statmod)
library(rvest)
library(UpSetR)

options(stringsAsFactors = FALSE)
set.seed(42)

`%notin%` <- Negate(`%in%`)

select   <- dplyr::select
rename   <- dplyr::rename
filter   <- dplyr::filter
distinct <- dplyr::distinct
mutate   <- dplyr::mutate

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_rect(colour="black", fill = NA, size = 1),
          axis.text = element_text(colour="black"),
          axis.ticks = element_line(colour="black"),
          axis.text.x = element_text(angle=0),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 7)))

pcutoff = 0.1
logPcutoff = -log10(pcutoff)
logFCcutoff = 0

setwd("/Users/tybottorff/git/Bottorff_CD127_2026/ABATE_BULK_ATAC/")

```

# Bring in metadata

```{r}

design <- read.csv("./saved_data/metadata_2.csv") %>%
  dplyr::mutate(donorId = as.character(donorId))

```

# Compare open chromatin between CD127+ Tex and CD127- Tex, export differential peaks for HOMER analyses

```{r}

peakFileName = "./saved_data/DP_CD127_pos_vs_DP_CD127_neg_diffbind_object.rds"
design_DP_CD127_pos_vs_DP_CD127_neg = design %>% filter(sort %in% c("TIGIT+KLRG1+\nCD57-CD127+", "TIGIT+KLRG1+\nCD57+CD127-", "TIGIT+KLRG1+\nCD57-CD127-PD1+")) %>% mutate(CD127_status = if_else(sort == "TIGIT+KLRG1+\nCD57-CD127+", "positive", "negative"))
# add a replicate index column (required by diffBind)
design_DP_CD127_pos_vs_DP_CD127_neg <- 
  design_DP_CD127_pos_vs_DP_CD127_neg %>%
  group_by(responder, sort) %>%
  mutate(replicate = row_number()) %>%
  ungroup() %>%
  # fix contrast
  mutate(sort = gsub("\\+", "pos", sort),
         sort = gsub("\\-", "neg", sort))
if (file.exists(peakFileName)) {
  consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg = readRDS(peakFileName)
} else {
  samples = data.frame(SampleID = design_DP_CD127_pos_vs_DP_CD127_neg$sample_name, 
                       Tissue = design_DP_CD127_pos_vs_DP_CD127_neg$sort,
                       Condition = design_DP_CD127_pos_vs_DP_CD127_neg$responder, 
                       Treatment = design_DP_CD127_pos_vs_DP_CD127_neg$CD127_status, 
                       Replicate = design_DP_CD127_pos_vs_DP_CD127_neg$replicate, 
                       Donor = design_DP_CD127_pos_vs_DP_CD127_neg$donorId,
                       bamReads = design_DP_CD127_pos_vs_DP_CD127_neg$bamFile, 
                       Peaks = design_DP_CD127_pos_vs_DP_CD127_neg$peakFiles, 
                       PeakCaller = "narrow")
  consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg = dba(sampleSheet=samples)
  summary(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg$binding[,3]-consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg$binding[,2])
  # this merged_peaks.bed file needs to be constructed from raw data files following scripts
  consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg = dba.count(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, peaks = "/Users/tybottorff/mounts/pipeline/Illumina/240327_VH00126_344_AACHGGFHV/Project_P576-1Processed_bri_240401/merged_peaks/all_merged_peaks.bed", summits = T)
  dba.plotPCA(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, attributes = c(DBA_TREATMENT, DBA_TISSUE, DBA_CONDITION))
  consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg = dba.contrast(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, categories=DBA_TREATMENT, block = c(DBA_REPLICATE, DBA_CONDITION, DBA_TISSUE), reorderMeta = list(Treatment = "positive"))
  consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg = dba.normalize(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, method=DBA_ALL_METHODS, library=DBA_LIBSIZE_FULL, normalize=DBA_NORM_LIB, background=FALSE)
  consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg = dba.analyze(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, method = DBA_ALL_METHODS)
  # PCA of differentially accessible regions
  dba.plotPCA(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, contrast = 1, attributes = c(DBA_TREATMENT, DBA_TISSUE, DBA_CONDITION))
  dba.show(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, bContrast=T)
  dba.plotVolcano(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, contrast=1)
  saveRDS(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, file = pathToDataFile)
}
annoData <- ChIPpeakAnno::toGRanges(EnsDb.Hsapiens.v86, feature="gene")
annoData = annoData[seqnames(annoData) %in% paste0("chr",c(1:23,"X","Y"))]
ucsc.levels <- paste0("chr",c(1:23,"X","Y"))
seqlevels(annoData) <- ucsc.levels
seqlevelsStyle(annoData) <- "NCBI"
consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg = dba.peakset(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, bRetrieve=T, DataType = DBA_DATA_FRAME)
filterChr = c(1:23, "X", "Y")
bed = consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg %>% dplyr::filter(START >= 0, CHR %in% filterChr) %>% dplyr::select(CHR, START, END)
consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg$peakID = 1:nrow(consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg)
consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g = GRanges(seqnames = consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg$CHR, 
                                 IRanges(start = consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg$START, 
                                         end = consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg$END), 
                                 strand = "*", 
                                 peakID = consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg$peakID)
seqlevelsStyle(consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g) = "UCSC"
seqlevels(consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g)
consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g = consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g[seqnames(consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g) %in% paste0("chr",c(1:23,"X","Y"))]
seqlevels(consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g) <- ucsc.levels
seqlevelsStyle(consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g) <- "NCBI"
seqlevels(consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g)
# first annotate the closest gene bodies (one to one)
geneBodies.g <- ChIPpeakAnno::annotatePeakInBatch(consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g, 
                                                  AnnotationData = annoData, 
                                                  output = "nearestLocation",
                                                  select = "first")
geneBodies.g = ChIPpeakAnno::addGeneIDs(geneBodies.g, orgAnn="org.Hs.eg.db", IDs2Add=c("symbol"))
nearestGeneBodies = data.frame(peakID = geneBodies.g$peakID, 
                               closestGene = geneBodies.g$symbol, 
                               distanceToGene = geneBodies.g$distancetoFeature, 
                               closestGeneENS = geneBodies.g$feature)
# next, annotate all promoters within +/- 10 000 bp (one to many)
nearPromoters.g = ChIPpeakAnno::annotatePeakInBatch(consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g, 
                                                    AnnotationData = annoData, 
                                                    output = "nearestBiDirectionalPromoters",
                                                    bindingRegion = c(-10000, 10000),
                                                    select = "all")
nearPromoters.g = ChIPpeakAnno::addGeneIDs(nearPromoters.g, orgAnn="org.Hs.eg.db", IDs2Add=c("symbol"))
nearPromoters = data.frame(peakID = nearPromoters.g$peakID, 
                           distanceToPromoter = nearPromoters.g$distanceToSite,
                           promoterSymbol = nearPromoters.g$gene_name,
                           promoterENS = nearPromoters.g$feature)
# all peaks that weren't annotated, we are now going to find, and then look for the one closest promoter within 100 000 bp (one to one)
missingPeakIDs = setdiff(consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g$peakID, nearPromoters.g$peakID)
missingPeaks.g = consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g[consensusPeaklist_DP_CD127_pos_vs_DP_CD127_neg.g$peakID %in% missingPeakIDs]
farPromoters.g = ChIPpeakAnno::annotatePeakInBatch(missingPeaks.g, 
                                                   AnnotationData = annoData, 
                                                   output = "nearestBiDirectionalPromoters",
                                                   bindingRegion = c(-100000, 100000),
                                                   select = "all")
farPromoters = data.frame(peakID = farPromoters.g$peakID,
                          distanceToPromoter = farPromoters.g$distanceToSite,
                          promoterSymbol = farPromoters.g$gene_name, 
                          promoterENS = farPromoters.g$feature)
promoters = rbind(nearPromoters, farPromoters)
promoters = dplyr::distinct(promoters)
# process contrasts
contrasts = dba.show(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, bContrast=T)
contrasts = dplyr::mutate(contrasts, name = paste0(Group, "_vs_", Group2))
dge_DP_CD127_pos_vs_DP_CD127_neg = c()
for (iContrast in 1:nrow(contrasts)) {
  dge_ = dba.report(consensusPeaks_DP_CD127_pos_vs_DP_CD127_neg, method=DBA_DESEQ2, contrast = iContrast, th = 1, DataType = DBA_DATA_FRAME)
  dge_$peakID = as.numeric(rownames(dge_))
  dge_ = merge(dge_, nearestGeneBodies, by = "peakID", all.x = TRUE, no.dups = FALSE)
  dge_ = left_join(dge_, promoters, by = "peakID", keep = FALSE) 
  dge_$peakPosition = paste0(dge_$Chr, ":", dge_$Start, "-", dge_$End)
  dge_ = dplyr::arrange(dge_, FDR)
  dge_DP_CD127_pos_vs_DP_CD127_neg[[iContrast]] = dge_
}
names(dge_DP_CD127_pos_vs_DP_CD127_neg) = contrasts$name
# Export differential peaks for HOMER analyses
for (contrast in names(dge_DP_CD127_pos_vs_DP_CD127_neg)) {
  dge_ = dge_DP_CD127_pos_vs_DP_CD127_neg[[contrast]]
  
  out = dge_ %>%
        dplyr::filter(FDR < 0.1) %>%
        dplyr::select(peakPosition, Fold, FDR, 
                      promoterSymbol, promoterENS, distanceToPromoter,
                      closestGene, closestGeneENS, distanceToGene
                      ) %>%
        dplyr::rename(logFC = Fold)
                      
  write.table(out, file =  "./saved_data/DP_CD127_pos_vs_DP_CD127_neg_differential_peaks_annotated.csv", quote=F, row.names=F)
  bed = dge_ %>% dplyr::filter(FDR < pcutoff) %>% dplyr::select(Chr, Start, End, Fold)
  bed = distinct(bed)
  # add 3 more cols needed by Homer
  bed$peakID = 1:nrow(bed)
  bed$blank = "1"
  bed$strand = "+"
  bed$Chr = paste("chr", bed$Chr, sep = "")
  # use consensus OCRs (open chromatin regions) from all non-naive CD8s as background for HOMER
  write.table(dplyr::select(bed, peakID, Chr, Start, End, strand), file = "./saved_data/DP_CD127_pos_vs_DP_CD127_neg_all_peaks.txt", sep="\t", quote=F, row.names=F, col.names=F)
}

```

# Read in HOMER output, plot enrichment of transcription factor motifs in open chromatin between Tex subsets (figure 6B, supplementary data 3)

```{r}

# look at both directions of a comparison and, for each motif, to take the side that's its more signfiicant on (sometimes end up with same motif called "enriched" on both sides of a comparison due to comparison being to background of all consensus peaks)
# look at both known and de novo motifs
known_motifs_dp_cd127_pos_vs_dp_pd1_pos_all_peaks_background <- read.table("./saved_data/DP_CD127_pos_vs_DP_PD1_pos_background_all_peaks_homer/knownResults.txt", header = TRUE, row.names = NULL) %>%
  dplyr::rename(pval = Name,
         perc_targets_with_motif = P.value.1,
         perc_background_with_motif = X.Benjamini.,
         num_targets_with_motif = Log,
         num_background_with_motif = q.value,
         full_motif_name = row.names,
         motif_seq = Motif,
         qval = P.value) %>%
  mutate(short_motif_name = str_split(full_motif_name, "/") %>% sapply(`[[`, 1)) %>%
  select(short_motif_name, pval, motif_seq, num_targets_with_motif, perc_targets_with_motif, num_background_with_motif, perc_background_with_motif) %>%
  mutate(source = "known")
de_novo_motifs_dp_cd127_pos_vs_dp_pd1_pos_all_peaks_background_html <- read_html("./saved_data/DP_CD127_pos_vs_DP_PD1_pos_background_all_peaks_homer/homerResults.html")
de_novo_motifs_dp_cd127_pos_vs_dp_pd1_pos_all_peaks_background_tables <- html_table(de_novo_motifs_dp_cd127_pos_vs_dp_pd1_pos_all_peaks_background_html, header = TRUE)
de_novo_motifs_dp_cd127_pos_vs_dp_pd1_pos_all_peaks_background <- de_novo_motifs_dp_cd127_pos_vs_dp_pd1_pos_all_peaks_background_tables[[1]] %>%
  dplyr::rename(pval = `P-value`,
         perc_targets_with_motif = `% of Targets`,
         perc_background_with_motif = `% of Background`,
         motif_seq = Motif,
         full_motif_name = `Best Match/Details`) %>%
  mutate(short_motif_name = str_split(full_motif_name, "/") %>% sapply(`[[`, 1)) %>%
  select(short_motif_name, pval, motif_seq, perc_targets_with_motif, perc_background_with_motif) %>%
  mutate(source = "de_novo")
motifs_dp_cd127_pos_vs_dp_pd1_pos_all_peaks_background <- known_motifs_dp_cd127_pos_vs_dp_pd1_pos_all_peaks_background %>%
  full_join(de_novo_motifs_dp_cd127_pos_vs_dp_pd1_pos_all_peaks_background) %>%
  mutate(enrichment = "DP CD127+")
known_motifs_dp_pd1_pos_vs_dp_cd127_pos_all_peaks_background <- read.table("./saved_data/DP_PD1_pos_vs_DP_CD127_pos_background_all_peaks_homer/knownResults.txt", header = TRUE, row.names = NULL) %>%
  dplyr::rename(pval = Name,
         perc_targets_with_motif = P.value.1,
         perc_background_with_motif = X.Benjamini.,
         num_targets_with_motif = Log,
         num_background_with_motif = q.value,
         full_motif_name = row.names,
         motif_seq = Motif,
         qval = P.value) %>%
  mutate(short_motif_name = str_split(full_motif_name, "/") %>% sapply(`[[`, 1)) %>%
  select(short_motif_name, pval, motif_seq, num_targets_with_motif, perc_targets_with_motif, num_background_with_motif, perc_background_with_motif) %>%
  mutate(source = "known")
de_novo_motifs_dp_pd1_pos_vs_dp_cd127_pos_all_peaks_background_html <- read_html("./saved_data/DP_PD1_pos_vs_DP_CD127_pos_background_all_peaks_homer/homerResults.html")
de_novo_motifs_dp_pd1_pos_vs_dp_cd127_pos_all_peaks_background_tables <- html_table(de_novo_motifs_dp_pd1_pos_vs_dp_cd127_pos_all_peaks_background_html, header = TRUE)
de_novo_motifs_dp_pd1_pos_vs_dp_cd127_pos_all_peaks_background <- de_novo_motifs_dp_pd1_pos_vs_dp_cd127_pos_all_peaks_background_tables[[1]] %>%
  dplyr::rename(pval = `P-value`,
         perc_targets_with_motif = `% of Targets`,
         perc_background_with_motif = `% of Background`,
         motif_seq = Motif,
         full_motif_name = `Best Match/Details`) %>%
  mutate(short_motif_name = str_split(full_motif_name, "/") %>% sapply(`[[`, 1)) %>%
  select(short_motif_name, pval, motif_seq, perc_targets_with_motif, perc_background_with_motif) %>%
  mutate(source = "de_novo")
motifs_dp_pd1_pos_vs_dp_cd127_pos_all_peaks_background <- known_motifs_dp_pd1_pos_vs_dp_cd127_pos_all_peaks_background %>%
  full_join(de_novo_motifs_dp_pd1_pos_vs_dp_cd127_pos_all_peaks_background) %>%
  mutate(enrichment = "DP PD-1+")
joined_dp_cd127_vs_dp_pd1_motifs <- motifs_dp_pd1_pos_vs_dp_cd127_pos_all_peaks_background %>%
  full_join(motifs_dp_cd127_pos_vs_dp_pd1_pos_all_peaks_background) %>%
  # take only 1 hit per motif (whatever side of contrast it was more strongly enriched)
  group_by(short_motif_name) %>%
  arrange(pval) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  # remove cols with data not present in all dfs
  select(-num_targets_with_motif, -num_background_with_motif) %>%
  # if perc_background_with_motifs = 0%, make it 0.01% to get FCs
  mutate(perc_background_with_motif = ifelse(perc_background_with_motif == "0.00%", "0.01%", perc_background_with_motif)) %>%
  mutate(fold_change = ifelse(enrichment == "DP CD127+", 
                            -log2(as.numeric(gsub("%", "", perc_targets_with_motif)) / as.numeric(gsub("%", "", perc_background_with_motif))),
                            log2(as.numeric(gsub("%", "", perc_targets_with_motif)) / as.numeric(gsub("%", "", perc_background_with_motif))))) %>%
  mutate(fold_change = if_else(perc_targets_with_motif == "0.00%", 0, fold_change)) %>%
  filter(abs(fold_change) > 0)
known_motifs_dp_cd127_pos_vs_dp_cd57_pos_all_peaks_background <- read.table("./saved_data/DP_CD127_pos_vs_DP_CD57_pos_background_all_peaks_homer/knownResults.txt", header = TRUE, row.names = NULL) %>%
  dplyr::rename(pval = Name,
         perc_targets_with_motif = P.value.1,
         perc_background_with_motif = X.Benjamini.,
         num_targets_with_motif = Log,
         num_background_with_motif = q.value,
         full_motif_name = row.names,
         motif_seq = Motif,
         qval = P.value) %>%
  mutate(short_motif_name = str_split(full_motif_name, "/") %>% sapply(`[[`, 1)) %>%
  select(short_motif_name, pval, motif_seq, num_targets_with_motif, perc_targets_with_motif, num_background_with_motif, perc_background_with_motif) %>%
  mutate(source = "known")
de_novo_motifs_dp_cd127_pos_vs_dp_cd57_pos_all_peaks_background_html <- read_html("./saved_data/DP_CD127_pos_vs_DP_CD57_pos_background_all_peaks_homer/homerResults.html")
de_novo_motifs_dp_cd127_pos_vs_dp_cd57_pos_all_peaks_background_tables <- html_table(de_novo_motifs_dp_cd127_pos_vs_dp_cd57_pos_all_peaks_background_html, header = TRUE)
de_novo_motifs_dp_cd127_pos_vs_dp_cd57_pos_all_peaks_background <- de_novo_motifs_dp_cd127_pos_vs_dp_cd57_pos_all_peaks_background_tables[[1]] %>%
  dplyr::rename(pval = `P-value`,
         perc_targets_with_motif = `% of Targets`,
         perc_background_with_motif = `% of Background`,
         motif_seq = Motif,
         full_motif_name = `Best Match/Details`) %>%
  mutate(short_motif_name = str_split(full_motif_name, "/") %>% sapply(`[[`, 1)) %>%
  select(short_motif_name, pval, motif_seq, perc_targets_with_motif, perc_background_with_motif) %>%
  mutate(source = "de_novo")
motifs_dp_cd127_pos_vs_dp_cd57_pos_all_peaks_background <- known_motifs_dp_cd127_pos_vs_dp_cd57_pos_all_peaks_background %>%
  full_join(de_novo_motifs_dp_cd127_pos_vs_dp_cd57_pos_all_peaks_background) %>%
  mutate(enrichment = "DP CD127+")
known_motifs_dp_cd57_pos_vs_dp_cd127_pos_all_peaks_background <- read.table("./saved_data/DP_CD57_pos_vs_DP_CD127_pos_background_all_peaks_homer/knownResults.txt", header = TRUE, row.names = NULL) %>%
  dplyr::rename(pval = Name,
         perc_targets_with_motif = P.value.1,
         perc_background_with_motif = X.Benjamini.,
         num_targets_with_motif = Log,
         num_background_with_motif = q.value,
         full_motif_name = row.names,
         motif_seq = Motif,
         qval = P.value) %>%
  mutate(short_motif_name = str_split(full_motif_name, "/") %>% sapply(`[[`, 1)) %>%
  select(short_motif_name, pval, motif_seq, num_targets_with_motif, perc_targets_with_motif, num_background_with_motif, perc_background_with_motif) %>%
  mutate(source = "known")
de_novo_motifs_dp_cd57_pos_vs_dp_cd127_pos_all_peaks_background_html <- read_html("./saved_data/DP_CD57_pos_vs_DP_CD127_pos_background_all_peaks_homer/homerResults.html")
de_novo_motifs_dp_cd57_pos_vs_dp_cd127_pos_all_peaks_background_tables <- html_table(de_novo_motifs_dp_cd57_pos_vs_dp_cd127_pos_all_peaks_background_html, header = TRUE)
de_novo_motifs_dp_cd57_pos_vs_dp_cd127_pos_all_peaks_background <- de_novo_motifs_dp_cd57_pos_vs_dp_cd127_pos_all_peaks_background_tables[[1]] %>%
  dplyr::rename(pval = `P-value`,
         perc_targets_with_motif = `% of Targets`,
         perc_background_with_motif = `% of Background`,
         motif_seq = Motif,
         full_motif_name = `Best Match/Details`) %>%
  mutate(short_motif_name = str_split(full_motif_name, "/") %>% sapply(`[[`, 1)) %>%
  select(short_motif_name, pval, motif_seq, perc_targets_with_motif, perc_background_with_motif) %>%
  mutate(source = "de_novo")
motifs_dp_cd57_pos_vs_dp_cd127_pos_all_peaks_background <- known_motifs_dp_cd57_pos_vs_dp_cd127_pos_all_peaks_background %>%
  full_join(de_novo_motifs_dp_cd57_pos_vs_dp_cd127_pos_all_peaks_background) %>%
  mutate(enrichment = "DP CD57+")
joined_dp_cd127_vs_dp_cd57_motifs <- motifs_dp_cd57_pos_vs_dp_cd127_pos_all_peaks_background %>%
  full_join(motifs_dp_cd127_pos_vs_dp_cd57_pos_all_peaks_background) %>%
  # take only 1 hit per motif (whatever side of contrast it was more strongly enriched)
  group_by(short_motif_name) %>%
  arrange(pval) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  # remove cols with data not present in all dfs
  select(-num_targets_with_motif, -num_background_with_motif) %>%
  # if perc_background_with_motifs = 0%, make it 0.01% to get FCs
  mutate(perc_background_with_motif = ifelse(perc_background_with_motif == "0.00%", "0.01%", perc_background_with_motif)) %>%
  mutate(fold_change = ifelse(enrichment == "DP CD127+", 
                            -log2(as.numeric(gsub("%", "", perc_targets_with_motif)) / as.numeric(gsub("%", "", perc_background_with_motif))),
                            log2(as.numeric(gsub("%", "", perc_targets_with_motif)) / as.numeric(gsub("%", "", perc_background_with_motif))))) %>%
  mutate(fold_change = if_else(perc_targets_with_motif == "0.00%", 0, fold_change)) %>%
  filter(abs(fold_change) > 0)
df1 <- joined_dp_cd127_vs_dp_cd57_motifs |>
  dplyr::mutate(comparison = "CD127+ Tex vs. CD57+ Tex")
df2 <- joined_dp_cd127_vs_dp_pd1_motifs |>
  dplyr::mutate(comparison = "CD127+ Tex vs. PD-1+ Tex")
plot_df <- dplyr::bind_rows(df1, df2) %>%
  mutate(
    enrichment = as.character(enrichment),
    enrichment = case_when(
      pval > 0.05 | abs(fold_change) < 1 ~ "NS",
      enrichment == "DP CD127+" ~ "CD127+ Tex",
      enrichment == "DP PD-1+" ~ "PD-1+ Tex",
      enrichment == "DP CD57+" ~ "CD57+ Tex"
      ),
    # clean motif names first
    short_motif_name_clean = sub("[\\(.].*$", "", short_motif_name) |> trimws(),
    source = case_when(
      source == "de_novo" ~ "de novo",
      source == "known" ~ "known"
    ))
labeled_df <- plot_df %>%
  filter(enrichment != "NS") %>%
  group_by(comparison, sign(fold_change)) %>%
  slice_min(pval, n = 5) %>%
  ungroup()
fig6B <- ggplot(plot_df, aes(
  x = fold_change,
  y = -log10(pval),
  color = enrichment
)) +
  geom_point(alpha = 0.75, size = 1.25) +
  geom_text_repel(
  data = labeled_df,
  aes(
    x = fold_change,
    y = -log10(pval),
    label = short_motif_name_clean,
    color = enrichment
  ),
  inherit.aes = FALSE,
  size = 4,
  box.padding = 0.22,
  point.padding = 0.2,
  max.overlaps = 9
  ) +
  scale_color_manual(values = c(
    "CD127+ Tex" = "#BF6E7A",
    "CD57+ Tex"  = "#302583",
    "PD-1+ Tex"  = "#37773C",
    "NS"        = "grey80"
  )) +
  labs(
    x = "Enrichment of motif in\ntarget over background (log2)",
    y = "-log10(p-value)",
    color = "Enrichment"
  ) +
  facet_wrap(~comparison) +
  theme(legend.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        strip.text = element_text(size = 9))
fig6B
ggsave("./figures/fig6B.pdf", fig6B, width = 6, height = 3)
write.csv(plot_df, "./saved_data/supp_data_3.csv", row.names = FALSE)

```

# Read in mitochondrial variant counts, plot quality control tests (supplementary figure 7)

```{r}

file_list <- list.files(path = "./saved_data/", pattern = "^merged_mt_.*\\.txt$", full.names = TRUE)
has_content <- function(file_path) {
  length(readLines(file_path)) > 1
}
filtered_file_list <- file_list[sapply(file_list, has_content)]
read_and_label <- function(file_path) {
  data <- read.table(file_path, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  data$file_name <- basename(file_path)
  data$libid <- str_extract(data$file_name, "lib[0-9]+")
  data$min_coverage <- as.numeric(str_extract(data$file_name, "(?<=coverage)\\d+"))
  data$p_value <- as.numeric(str_replace(str_extract(data$file_name, "(?<=pvalue)[0-9]+_[0-9]+"), "_", "."))
  data$min_var_freq <- as.numeric(str_replace(str_extract(data$file_name, "(?<=minvarfreq)[0-9]+_[0-9]+"), "_", "."))
  return(data)
}
labeled_data_list <- lapply(filtered_file_list, read_and_label)
combined_data <- do.call(rbind, labeled_data_list)
# get these downstream but using them now
vars_present_in_all_sorts_in_donors <- read.csv("./saved_data/vars_in_all_sorts.csv") %>%
  select(var_id) %>%
  rename(donor_position_allele_id = var_id)
public_vars <- read.csv("./saved_data/public_vars.csv") %>%
  select(var_id) %>%
  rename(position_allele_id = var_id)
metadata <- read.csv("./saved_data/metadata.csv") %>%
  # fix donorId...
  mutate(donorId = paste("00", donorId, sep = ""),
         donorId = as.character(donorId))
mt_var_data <- combined_data %>%
  left_join(design %>% dplyr::select(libid, donorId, sort, responder), by = "libid") %>%
  left_join(metadata, by = c("donorId", "responder")) %>%
  # many to many warning in left_join expected
  dplyr::rename(Chromosome = Chrom,
         Reference_allele = Ref,
         Consensus_genotype = Cons,
         Reads_supporting_ref_allele = Reads1,
         Reads_supporting_var_allele = Reads2,
         ref_allele_strands = Strands1,
         var_allele_strands = Strands2,
         avg_base_qual_of_ref_supporting_read_bases = Qual1,
         avg_base_qual_of_var_supporting_read_bases = Qual2,
         avg_map_qual_of_ref_reads = MapQual1,
         avg_map_qual_of_var_reads = MapQual2,
         ref_supporting_reads_on_plus_strand = Reads1Plus,
         ref_supporting_reads_on_minus_strand = Reads1Minus,
         var_supporting_reads_on_plus_strand = Reads2Plus,
         var_supporting_reads_on_minus_strand = Reads2Minus,
         most_freq_var_allele = VarAllele,
         Var_perc = VarFreq) %>%
  # apply filtering steps to avoid false positive vars, as done in https://elifesciences.org/articles/45105#s4
  dplyr::filter(!(Position %in% c(302:316, 513:525, 3105:3110)),
         var_supporting_reads_on_plus_strand > 2,
         var_supporting_reads_on_minus_strand > 2,
         ref_supporting_reads_on_plus_strand > 2,
         ref_supporting_reads_on_minus_strand > 2,
         # strand balance check
         var_supporting_reads_on_plus_strand / (var_supporting_reads_on_plus_strand + var_supporting_reads_on_minus_strand) > 0.3,
         var_supporting_reads_on_plus_strand / (var_supporting_reads_on_plus_strand + var_supporting_reads_on_minus_strand) < 0.7,
         ref_supporting_reads_on_plus_strand / (ref_supporting_reads_on_plus_strand + ref_supporting_reads_on_minus_strand) > 0.3,
         ref_supporting_reads_on_plus_strand / (ref_supporting_reads_on_plus_strand + ref_supporting_reads_on_minus_strand) < 0.7) %>%
  mutate(Var_perc = as.numeric(gsub("%", "", Var_perc)),
         sort = case_when(
           sort == "TIGIT-\nKLRG1-" ~ "DN",
           sort == "(TIGIT+KLRG1+)-\nCD127+" ~ "non-exhausted CD127+",
           sort == "TIGIT+KLRG1+\nCD57-CD127+" ~ "DP CD127+",
           sort == "TIGIT+KLRG1+\nCD57+CD127-" ~ "DP CD57+",
           sort == "TIGIT+KLRG1+\nCD57-CD127-PD1+" ~ "DP PD-1+")) %>%
  # remove vars present in all donors in a sort
  mutate(donor_position_allele_id = paste(donorId, Position, most_freq_var_allele, sep = ", "),
         position_allele_id = paste(Position, most_freq_var_allele, sep = ", ")) %>%
  filter(!(donor_position_allele_id %in% vars_present_in_all_sorts_in_donors$donor_position_allele_id),
         !(position_allele_id %in% public_vars$position_allele_id))
hv1_start <- 16024
hv1_end <- 16383
hv2_start <- 57
hv2_end <- 372
supp_fig_6A <- ggplot(data = mt_var_data, aes(x = Position)) +
  geom_histogram(binwidth = 250) +
  labs(x = "Position", y = "Number of variants") +
  geom_vline(xintercept = c(hv1_start, hv1_end), color = "red", linetype = "dashed") +
  geom_vline(xintercept = c(hv2_start, hv2_end), color = "blue", linetype = "dashed")
supp_fig_6A
ggsave("./figures/supp_fig_6A.pdf", supp_fig_6A, width = 5, height = 5)
mt_var_data$arbitrary_id <- as.numeric(factor(mt_var_data$donorId, levels = unique(mt_var_data$donorId)))
upset_mt_var_data_2 <- mt_var_data %>%
  group_by(Position, most_freq_var_allele) %>%
  summarize(donors = list(unique(arbitrary_id)), .groups = 'drop') %>%
  unnest(donors) %>%
  mutate(value = 1) %>%
  pivot_wider(names_from = donors, values_from = value, values_fill = list(value = 0))
supp_fig_6B <- upset(as.data.frame(upset_mt_var_data_2), sets = setdiff(colnames(upset_mt_var_data_2), c("sort", "Position", "most_freq_var_allele")), sets.bar.color = "#56B4E9", sets.x.label = "MT SNV counts", mainbar.y.label = "MT SNV sharing across donors", point.size = 1.5, show.numbers = FALSE, text.scale = 1)
supp_fig_6B
pdf("./figures/supp_fig_6B.pdf", height = 4.5, width = 5)
print(supp_fig_6B)
invisible(dev.off())
# this is where public variants CSV came from
public_vars <- upset_mt_var_data_2 %>%
  filter(`1` + `2` + `3` + `4` + `5` + `6` + `7` + `8` + `9` + `10` > 1) %>%
  mutate(var_id = paste(Position, most_freq_var_allele, sep = ", "))
#write.csv(public_vars, "./saved_data/public_vars.csv")
# this is where variants in all cell populations CSV came from
upset_mt_var_data <- mt_var_data %>%
  group_by(donorId, Position, most_freq_var_allele) %>%
  summarize(sorts = list(unique(sort)), .groups = 'drop') %>%
  unnest(sorts) %>%
  mutate(value = 1) %>%
  pivot_wider(names_from = sorts, values_from = value, values_fill = list(value = 0))
vars_present_in_all_sorts_in_donors <- upset_mt_var_data %>%
  filter(`DP CD127+` + `non-exhausted CD127+` + `DN` + `DP CD57+` + `DP PD-1+` == 5) %>%
  mutate(var_id = paste(donorId, Position, most_freq_var_allele, sep = ", "))
#write.csv(vars_present_in_all_sorts_in_donors, "./saved_data/vars_in_all_sorts.csv")

```

# Plot mitochondrial variant counts for CD8 subsets (figure 7C)

```{r}

count_data <- mt_var_data %>%
  mutate(var_perc_category = case_when(
    Var_perc <= 25 ~ "< 5%",
    Var_perc > 5 & Var_perc <= 99 ~ "25-99%",
    Var_perc > 99 ~ "> 99%")) %>%
    mutate(var_perc_category = factor(var_perc_category, levels = c("< 5%", "5-25%", "25-99%", "> 99%"))) %>%
  group_by(sort, donorId, responder, var_perc_category) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(key = "A")
count_data_all <- count_data %>%
  group_by(sort, donorId, responder) %>%
  summarize(count = sum(count), var_perc_category = "All", .groups = "drop") %>%
  mutate(key = "B")
summed_data <- bind_rows(count_data, count_data_all) %>%
  mutate(var_perc_category = factor(var_perc_category, levels = c("All", "< 5%", "5-25%", "25-99%", "> 99%")))
summed_data$sort <- factor(summed_data$sort, levels = c("non-exhausted CD127+", "DN", "DP CD127+", "DP PD-1+", "DP CD57+"))
custom_labeller <- function(var_perc_category) {
  paste("Variant freq:", var_perc_category)
}
custom_labels <- c(
  "< 5%" = "Rare variants",
  "> 99%" = "Common variants"
)
summed_data <- summed_data %>%
  mutate(
    sort = case_when(
      sort == "DN" ~ "TIGIT-KLRG1-",
      sort == "DP CD127+" ~ "CD127+ Tex",
      sort == "DP PD-1+" ~ "PD-1+ Tex",
      sort == "DP CD57+" ~ "CD57+ Tex",
      TRUE ~ sort
    ),
    sort = factor(
      sort,
      levels = c(
        "non-exhausted CD127+",
        "TIGIT-KLRG1-",
        "CD127+ Tex",
        "PD-1+ Tex",
        "CD57+ Tex"
      )
    )
  )
colors <- c("#9E4C95", "#7D2B54", "#BF6E7A", "#37773C", "#302583")
names(colors) <- c("TIGIT-KLRG1-", "non-exhausted CD127+", "CD127+ Tex", "PD-1+ Tex", "CD57+ Tex")
fig7C <- ggplot(summed_data %>% dplyr::filter(var_perc_category %in% c("< 5%", "> 99%")), aes(x = sort, y = count, fill = sort)) +
  geom_boxplot(outlier.shape = NA) +
  geom_beeswarm(size = 0.25, alpha = 0.5) +
  geom_line(aes(group = donorId), color = "grey", alpha = 0.5, size = 0.5) +
  facet_wrap(~ var_perc_category, ncol = 2, labeller = labeller(var_perc_category = custom_labels)) +
  theme(axis.text.x = element_text(size = 9, angle = 45, hjust = 1), strip.text = element_text(size = 10), axis.title.y = element_text(size = 12), axis.text.y = element_text(size = 10),
        legend.position = "none") +
  labs(x = "", y = "MT sequence\nvariant count", fill = "Population") +
  scale_fill_manual(values = colors) +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE))
fig7C
ggsave("./figures/fig7C.pdf", plot = fig7C, width = 3.5, height = 3.5)
perform_wilcox_tests <- function(data) {
  # Get unique cell sorts
  sorts <- unique(data$sort)
  
  # Generate pairwise combinations, keeping only one direction
  pairwise_combos <- combn(sorts, 2, simplify = FALSE) %>%
    map_df(~ tibble(Cell_Sort_1 = .x[1], Cell_Sort_2 = .x[2]))
  
  # Perform tests
  results <- pairwise_combos %>%
    rowwise() %>%
    mutate(p_value = {
      pair_data <- data %>% filter(sort %in% c(Cell_Sort_1, Cell_Sort_2))
      group1 <- pair_data %>% filter(sort == Cell_Sort_1) %>% pull(count)
      group2 <- pair_data %>% filter(sort == Cell_Sort_2) %>% pull(count)
      wilcox.test(group1, group2, paired = FALSE, exact = FALSE)$p.value
    }) %>%
    ungroup()
  
  return(results)
}
p_values_summary <- summed_data %>%
  group_by(var_perc_category) %>%
  group_modify(~ {
    res <- perform_wilcox_tests(.x)
    res %>%
      mutate(p_adjusted = p.adjust(p_value, method = "BH"))
  }) %>%
  ungroup() %>%
  arrange(p_adjusted)

```

# Plot peak scores for open chromatin near cytotoxicity genes for all non-naive CD8s (supplementary figure 8A)

```{r}

peakFileName = "./saved_data/grouped_diffbind_object.rds"
# add a replicate index column (required by diffBind)
design_grouped <- design %>%
  group_by(responder, sort) %>%
  mutate(replicate = row_number()) %>%
  ungroup() %>%
  mutate(responder = if_else(responder == "Yes", "Responder", "Non-responder"),
         sort = case_when(
           sort == "TIGIT-\nKLRG1-" ~ "DN",
           sort == "(TIGIT+KLRG1+)-\nCD127+" ~ "non-exhausted CD127+",
           sort == "TIGIT+KLRG1+\nCD57-CD127+" ~ "DP CD127+",
           sort == "TIGIT+KLRG1+\nCD57+CD127-" ~ "DP CD57+",
           sort == "TIGIT+KLRG1+\nCD57-CD127-PD1+" ~ "DP PD-1+"))
if (file.exists(peakFileName)) {
  consensusPeaks_grouped = readRDS(peakFileName)
} else {
  samples = data.frame(SampleID = design_grouped$sample_name, 
                       Tissue = design_grouped$sort,
                       Condition = design_grouped$responder, 
                       Treatment = design_grouped$responder, 
                       Replicate = design_grouped$replicate, 
                       bamReads = design_grouped$bamFile, 
                       Peaks = design_grouped$peakFiles, 
                       PeakCaller = "narrow")
  consensusPeaks_grouped = dba(sampleSheet=samples)
  consensusPeaks_grouped = dba.count(consensusPeaks_grouped, peaks = "/Users/tybottorff/mounts/240327_VH00126_344_AACHGGFHV/Project_P576-1Processed_bri_240401/merged_peaks/all_merged_peaks.bed", summits = T)
  consensusPeaks_grouped = dba.contrast(consensusPeaks_grouped, categories=DBA_TISSUE, block = c(DBA_REPLICATE, DBA_CONDITION))
  consensusPeaks_grouped = dba.normalize(consensusPeaks_grouped, method=DBA_ALL_METHODS, library=DBA_LIBSIZE_FULL, normalize=DBA_NORM_LIB, background=FALSE)
  consensusPeaks_grouped = dba.analyze(consensusPeaks_grouped, method = DBA_ALL_METHODS)
  pca_plot <- dba.plotPCA(consensusPeaks_grouped, attributes = c(DBA_TISSUE, DBA_CONDITION), vColors = c("skyblue", "black", "yellow", "darkblue", "magenta", "orange", "darkgreen", "red", "grey", "lightgreen"))
  pdf(plotDir("P576-1_cell_sort_and_responder_pca.pdf"))
  print(pca_plot)
  dev.off()
  pca_plot <- dba.plotPCA(consensusPeaks_grouped, attributes = c(DBA_TISSUE), vColors = c("#440154FF", "#3B528BFF", "#21908CFF", "#FDE725FF", "#5DC863FF"))
  pdf(plotDir("P576-1_cell_sort_pca.pdf"))
  print(pca_plot)
  dev.off()
  consensusPeaks_grouped$config$condition <- "Responder status"
  consensusPeaks_grouped$config$replicate <- "Donor"
  consensusPeaks_grouped$config$tissue <- "Cell population"
  consensusPeaks_grouped$class[DBA_ID,] <- ""
dba.plotHeatmap(consensusPeaks_grouped, correlations = FALSE, scale = "row", ColAttributes = c(DBA_TISSUE, DBA_CONDITION, DBA_REPLICATE), colSideCols = list(DBA_TISSUE = viridis(5), DBA_CONDITION = c("black", "grey"), DBA_REPLICATE = c("#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", "#A6D854", "#FFD92F")), margin = 2, add.expr = {
                   # Add legend for DBA_TISSUE
                   legend("right", legend = unique(consensusPeaks_grouped$samples$Tissue), 
                          fill = viridis(5),
                          title = "Cell population")
                 })
  dev.off()
  dba.show(consensusPeaks_grouped, bContrast=T)
  saveRDS(consensusPeaks_grouped, file = pathToDataFile)
}

# re-run steps outside diffbind
peaks_list <- consensusPeaks_grouped$peaks
count_matrix <- as.data.frame(peaks_list[[1]]$Reads)
for (i in 2:length(peaks_list)) {
  sample_data <- peaks_list[[i]]
  counts <- sample_data$Reads
  count_matrix <- cbind(count_matrix, counts)
}
tissue <- consensusPeaks_grouped$samples$Tissue
responder_status <- consensusPeaks_grouped$samples$Condition
colnames(count_matrix) <- paste(consensusPeaks_grouped$samples$Tissue, consensusPeaks_grouped$samples$Condition, consensusPeaks_grouped$samples$Replicate, sep = "_")
# normalize by library size
library_sizes <- colSums(count_matrix)
count_matrix <- sweep(count_matrix, 2, library_sizes, FUN = "/")
# regress out effect of responder status (minimal) in comparing cell subsets
regress_out_responder_status <- function(feature, responder_status) {
  model <- lm(feature ~ responder_status)
  residuals(model)
}
regress_out_tissue <- function(feature, tissue) {
  model <- lm(feature ~ tissue)
  residuals(model)
}
features <- t(count_matrix)
R_NR_adjusted_features <- sapply(1:ncol(features), function(i) {
  feature <- features[, i]
  regress_out_responder_status(feature, responder_status)
})
R_NR_adjusted_features_df <- as.data.frame(R_NR_adjusted_features)
tissue_adjusted_features <- sapply(1:ncol(features), function(i) {
  feature <- features[, i]
  regress_out_tissue(feature, tissue)
})
tissue_adjusted_features_df <- as.data.frame(tissue_adjusted_features)
# log2 adjust
tissue_adjusted_features_df_log2 <- log2(tissue_adjusted_features_df - min(tissue_adjusted_features_df) + 1)
pca_result_tissue_adjusted <- prcomp((tissue_adjusted_features_df_log2), scale. = TRUE)
pca_scores <- data.frame(pca_result_tissue_adjusted$x)
explained_variance <- summary(pca_result_tissue_adjusted)$importance[2, 1:3] * 100
pca_scores$responder_status <- consensusPeaks_grouped$samples$Condition
# log2 adjust
R_NR_adjusted_features_df_log2 <- log2(R_NR_adjusted_features_df - min(R_NR_adjusted_features_df) + 1)
pca_result_R_NR_adjusted <- prcomp((R_NR_adjusted_features_df_log2), scale. = TRUE)
pca_scores <- data.frame(pca_result_R_NR_adjusted$x)
explained_variance <- summary(pca_result_R_NR_adjusted)$importance[2, 1:2] * 100
pca_scores$Tissue <- consensusPeaks_grouped$samples$Tissue
pca_scores$Tissue <- factor(consensusPeaks_grouped$samples$Tissue, levels = c("DN", "non-exhausted CD127+", "DP CD127+", "DP PD-1+", "DP CD57+"))
tissue_colors <- viridis(5)
ggplot(pca_scores, aes(x = PC1, y = PC2, color = Tissue)) +
  geom_point(size = 2.5) +
  stat_ellipse(aes(group = Tissue), level = 0.95) +
  scale_color_manual(values = tissue_colors) +
  labs(x = paste("PC1 (", round(explained_variance[1], 2), "%)", sep = ""), y = paste("PC2 (", round(explained_variance[2], 2), "%)", sep = ""),
       color = "CD8 subpopulation") +
  theme(legend.text = element_text(size = 10))
compVar <- data.frame(PC1 = pca_scores$PC1,
                      PC2 = pca_scores$PC2,
                      PC3 = pca_scores$PC3,
                      group = pca_scores$Tissue)
annoData <- ChIPpeakAnno::toGRanges(EnsDb.Hsapiens.v86, feature="gene")
annoData = annoData[seqnames(annoData) %in% paste0("chr",c(1:23,"X","Y"))]
ucsc.levels <- paste0("chr",c(1:23,"X","Y"))
seqlevels(annoData) <- ucsc.levels
seqlevelsStyle(annoData) <- "NCBI"
consensusPeaklist_grouped = dba.peakset(consensusPeaks_grouped, bRetrieve=T, DataType = DBA_DATA_FRAME)
filterChr = c(1:23, "X", "Y")
bed = consensusPeaklist_grouped %>% dplyr::filter(START >= 0, CHR %in% filterChr) %>% dplyr::select(CHR, START, END)
consensusPeaklist_grouped$peakID = 1:nrow(consensusPeaklist_grouped)
consensusPeaklist_grouped.g = GRanges(seqnames = consensusPeaklist_grouped$CHR, 
                                 IRanges(start = consensusPeaklist_grouped$START, 
                                         end = consensusPeaklist_grouped$END), 
                                 strand = "*", 
                                 peakID = consensusPeaklist_grouped$peakID)
seqlevelsStyle(consensusPeaklist_grouped.g) = "UCSC"
seqlevels(consensusPeaklist_grouped.g)
consensusPeaklist_grouped.g = consensusPeaklist_grouped.g[seqnames(consensusPeaklist_grouped.g) %in% paste0("chr",c(1:23,"X","Y"))]
seqlevels(consensusPeaklist_grouped.g) <- ucsc.levels
seqlevelsStyle(consensusPeaklist_grouped.g) <- "NCBI"
seqlevels(consensusPeaklist_grouped.g)
# first annotate the closest gene bodies (one to one)
geneBodies.g <- ChIPpeakAnno::annotatePeakInBatch(consensusPeaklist_grouped.g, 
                                                  AnnotationData = annoData, 
                                                  output = "nearestLocation",
                                                  select = "first")
geneBodies.g = ChIPpeakAnno::addGeneIDs(geneBodies.g, orgAnn="org.Hs.eg.db", IDs2Add=c("symbol"))
nearestGeneBodies = data.frame(peakID = geneBodies.g$peakID, 
                               closestGene = geneBodies.g$symbol, 
                               distanceToGene = geneBodies.g$distancetoFeature, 
                               closestGeneENS = geneBodies.g$feature)
# next annotate all promoters within +/- 10 000 bp (one to many)
nearPromoters.g = ChIPpeakAnno::annotatePeakInBatch(consensusPeaklist_grouped.g, 
                                                    AnnotationData = annoData, 
                                                    output = "nearestBiDirectionalPromoters",
                                                    bindingRegion = c(-10000, 10000),
                                                    select = "first")
nearPromoters.g = ChIPpeakAnno::addGeneIDs(nearPromoters.g, orgAnn="org.Hs.eg.db", IDs2Add=c("symbol"))
nearPromoters = data.frame(peakID = nearPromoters.g$peakID, 
                           distanceToPromoter = nearPromoters.g$distanceToSite,
                           promoterSymbol = nearPromoters.g$gene_name,
                           promoterENS = nearPromoters.g$feature)
# all peaks that weren't annotated, we are now going to find, and then look for the one closest promoter within 100 000 bp (one to one)
missingPeakIDs = setdiff(consensusPeaklist_grouped.g$peakID, nearPromoters.g$peakID)
missingPeaks.g = consensusPeaklist_grouped.g[consensusPeaklist_grouped.g$peakID %in% missingPeakIDs]
farPromoters.g = ChIPpeakAnno::annotatePeakInBatch(missingPeaks.g, 
                                                   AnnotationData = annoData, 
                                                   output = "nearestBiDirectionalPromoters",
                                                   bindingRegion = c(-100000, 100000),
                                                   select = "first")
farPromoters = data.frame(peakID = farPromoters.g$peakID,
                          distanceToPromoter = farPromoters.g$distanceToSite,
                          promoterSymbol = farPromoters.g$gene_name, 
                          promoterENS = farPromoters.g$feature)
promoters = rbind(nearPromoters, farPromoters)
promoters = dplyr::distinct(promoters)
annotation_info <- nearestGeneBodies %>%
  full_join(nearPromoters, by = "peakID") %>%
  full_join(farPromoters, by = c("peakID", "distanceToPromoter", "promoterSymbol", "promoterENS"))
contrasts = dba.show(consensusPeaks_grouped, bContrast=T)
contrasts = dplyr::mutate(contrasts, name = paste0(Group, "_vs_", Group2))
dge_grouped = c()
for (iContrast in 1:nrow(contrasts)) {
  dge_ = dba.report(consensusPeaks_grouped, method=DBA_DESEQ2, contrast = iContrast, th = 1, DataType = DBA_DATA_FRAME)
  dge_$peakID = as.numeric(rownames(dge_))
  dge_ = merge(dge_, nearestGeneBodies, by = "peakID", all.x = TRUE, no.dups = FALSE)
  dge_ = left_join(dge_, promoters, by = "peakID", keep = FALSE) 
  dge_$peakPosition = paste0(dge_$Chr, ":", dge_$Start, "-", dge_$End)
  dge_ = dplyr::arrange(dge_, FDR)
  dge_grouped[[iContrast]] = dge_
}
names(dge_grouped) = contrasts$name
col_fun = colorRamp2(c(-4, 0, 4), c("#0033a0", "white", "#FE5000"))
reorder = order(design_grouped$sort, design_grouped$libid)
design_grouped_ <- design_grouped[reorder, ]
countsConsensusPeaks = as.matrix(consensusPeaklist_grouped[ , 4:(ncol(consensusPeaklist_grouped)-1)])
countsConsensusPeaks_ = countsConsensusPeaks[ , reorder]
design_grouped$sort <- ifelse(is.na(design_grouped$sort), "DN", design_grouped$sort)
design_grouped_$sort <- ifelse(is.na(design_grouped_$sort), "DN", design_grouped_$sort)
names(dge_grouped) <- if_else(
  str_starts(names(dge_grouped), "_"),
  paste0("DN", names(dge_grouped)),
  names(dge_grouped)
)
desired_order <- c("DN", "non-exhausted CD127+", "DP CD127+", "DP PD-1+", "DP CD57+")
design_grouped_$sort <- factor(design_grouped_$sort, levels = desired_order)
order_index <- order(design_grouped_$sort)
colors.sort = c("#9E4C95", "#7D2B54", "#BF6E7A", "#37773C", "#302583")
names(colors.sort) <- desired_order
cytotoxicity_genes <- c("GZMA", "GZMB", "GZMH", "GZMM", "PRF1", "NCR1", "KLRF1", "NKG7", "IFNG", "KLRC1", "KLR", "GNLY", "CST7", "FGFBP2")
for (contrast in names(dge_grouped)[1]) {
  dge_ = dge_grouped[[contrast]]
  sigPeaks = dge_ %>% filter(peakID %in% annotation_info$peakID) %>% group_by(peakID) %>%
  # take closest promoter for multiples annotated
  arrange(distanceToPromoter) %>%
  dplyr::slice(1) %>%
  ungroup()
  sigPeaks$geneSet <- case_when(
    sigPeaks$closestGene %in% cytotoxicity_genes | sigPeaks$promoterSymbol %in% cytotoxicity_genes ~ "Cytotoxicity",
    TRUE ~ "Other")
  sigPeaks <- sigPeaks %>%
    filter(geneSet == "Cytotoxicity")
  gene_set_labels <- data.frame(
    geneSet = sigPeaks$geneSet)
  sigPeaksNames = paste(sigPeaks$closestGene, sigPeaks$promoterSymbol, sep = ", ")
  sigPeaksIDs = sigPeaks$peakID
  counts_ = countsConsensusPeaks_[rownames(countsConsensusPeaks_) %in% sigPeaksIDs, ]
  rownames(counts_) = sigPeaksNames
  tp = t(scale(t(log2(1 + counts_))))
  tp = tp[, order_index]
  ha <- HeatmapAnnotation(df = list(group = design_grouped_$sort[order_index]), 
                          col = list(group = colors.sort)
                          )
  supp_fig_8A = Heatmap(tp, show_column_names = FALSE, show_row_names = TRUE, 
               cluster_columns = FALSE, name = "scaled\nlog2\npeak score", 
               top_annotation = ha, col = col_fun, cluster_rows = FALSE, show_row_dend = TRUE, column_title = "", row_title = "Cytotoxicity OCRs", row_names_gp = gpar(fontsize = 10), row_title_gp = gpar(fontsize = 20), heatmap_legend_param = list(title = "scaled\nlog2\npeak score", title_gp = gpar(fontsize = 20), labels_gp = gpar(fontsize = 20)), row_split = gene_set_labels$geneSet)
  plot(supp_fig_8A)
  pdf("./figures/supp_fig_8A.pdf", height = 7, width = 7.5)
  print(supp_fig_8A)
  invisible(dev.off())
}

```